---
title: "Fisher DNA + Cameras — State-Space Prep, Occasions & Detection Histories"
author: "Prepared for new analysts"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
  word_document: default
---

> **Purpose**: Read cleaned **DNA** (hair snagger) and **camera** data, apply project-specific rules, 
> assign **state spaces**, construct **occasion-based** structures for DNA (\`Y.s\`) and cameras (\`O.o\`, \`O.o2\`), and 
> save a workspace for downstream SCR / integrated models.
>
> **How to use?**  Update the **Parameters** (paths/season) and knit.

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8, fig.height = 6)
options(stringsAsFactors = FALSE, scipen = 999)
```

## Packages
Installs any missing packages (CRAN) and loads them.

```{r packages}
pkgs <- c(
  "sf","rmapshaper","dplyr","readr","lubridate","stringr","tidyr",
  "purrr","ggplot2","camtrapR","data.table","plyr","scales", "cowplot"
)
missing <- setdiff(pkgs, rownames(installed.packages()))
if (length(missing)) install.packages(missing, repos = "https://cloud.r-project.org")
invisible(lapply(pkgs, library, character.only = TRUE))
```

## Parameters — **Edit these**

```{r parameters}
species_of_interest <- "Pekania pennanti"
season              <- "all"     # "winter" or "all"
crs_wgs84           <- 4326
crs_albers          <- 3005          # BC Albers (meters)
buffer_m            <- 17000         # Columbian AOI buffer
coord_scale_km      <- 1000          # meters -> km
occasion_length     <- 14            # days for detectionHistory
tz_local            <- "America/Vancouver"
preserve_zero_projects <- TRUE

root       <- "//Sfp.idir.bcgov/s140/S40203/Ecosystems/Conservation Science/Species/Mesocarnivores/Projects/MMP/2.Data"
local_root <- "/2. GIS"

paths <- list(
  shp_meso_grid      = file.path(local_root, "BC_meso_grid.shp"),
  shp_columbian_grid = file.path(local_root, "BC_meso_grid_columbian.shp"),
  gdb_subpops        = file.path(root, "9. GIS/BC_Fisher_populations_2024_edited.gdb"),
  gdb_layer_subpops  = "Subpopulations",
  dna_clean          = file.path(root, "1. MDB-Collation/5. Master data cleaned/Species specific/Fisher/DNA_fisher_clean_states.csv"),
  cams_all           = file.path(root, "1. MDB-Collation/5. Master data cleaned/Species specific/Fisher/camera_deployments_singledeployment_2026-02-24.csv"),
  cams_winter        = file.path(root, "1. MDB-Collation/5. Master data cleaned/Species specific/Fisher/camera_deployments_singledeployment_winter2026-03-02.csv"),
  det_all            = file.path(root, "1. MDB-Collation/5. Master data cleaned/Species specific/Fisher/fisher_detections_singledeployment_2026-02-24.csv"),
  det_winter         = file.path(root, "1. MDB-Collation/5. Master data cleaned/Species specific/Fisher/fisher_detections_singledeployment_winter2026-03-02.csv"),
  out_dir            = local_root,
  save_rdata_winter  = file.path("./3. Outputs/Process_Fisher_camera-DNA_winter-only_Feb.RData"),
  save_rdata_all     = file.path("./3. Outputs/Process_Fisher_camera-DNA_allseasons_March.RData")
)
```

## Helper functions
Reusable utilities for sf conversion, project rules, state assignment, scaling, and bounds.

```{r helpers}
make_sf_points <- function(df, lon, lat, crs_in = 4326, crs_out = 3005) {
  df_sf <- sf::st_as_sf(df, coords = c(lon, lat), crs = crs_in, remove = FALSE) |>
    sf::st_transform(crs_out)
  coords <- sf::st_coordinates(df_sf)
  dplyr::mutate(df_sf, UTM_X = coords[,1], UTM_Y = coords[,2])
}

apply_project_rules <- function(df, rules_tbl) {
  out <- df
  for (i in seq_len(nrow(rules_tbl))) {
    pat <- rules_tbl$project_match[i]
    fun <- rules_tbl$rename_fun[[i]]
    idx <- which(out$Project_name == pat)
    if (length(idx)) out[idx, ] <- fun(out[idx, ])
  }
  out
}

assign_states <- function(df, state_map_tbl) {
  dplyr::left_join(df, state_map_tbl, by = "Project_name")
}

```

## Read spatial AOI and subpopulations
- Read and buffer the **Columbian** grid (17 km).
- Read **subpopulations** for context plotting.

```{r aoi}

grid_columbian <- sf::st_read(paste0(".",paths$shp_columbian_grid), quiet = TRUE) |>
  sf::st_transform(crs_albers)

columbian_area <- grid_columbian |>
  sf::st_union() |>
  rmapshaper::ms_simplify(keep = 0.01, keep_shapes = FALSE) |>
  sf::st_buffer(buffer_m)

subpopulations <- sf::st_read(paths$gdb_subpops, layer = paths$gdb_layer_subpops, quiet = TRUE) |>
  sf::st_transform(crs_albers) |>
  rmapshaper::ms_simplify(keep = 0.1, keep_shapes = FALSE)
```

## Read **DNA** cleaned data (already processed in prior script)
- Uses the **cleaned** CSV with state (if present) and coordinates.

```{r dna}
DNA <- readr::read_csv(paths$dna_clean, show_col_types = FALSE)
stopifnot(all(c("Longitude_DD","Latitude_DD") %in% names(DNA)))
DNA <- make_sf_points(DNA, lon = "Longitude_DD", lat = "Latitude_DD",
                      crs_in = crs_wgs84, crs_out = crs_albers)
```

## Read **camera deployments** & **detections**
Select inputs by `season`. Convert to sf and attach Albers coordinates.

```{r cams-read}
cams_file <- if (season == "winter") paths$cams_winter else paths$cams_all
det_file  <- if (season == "winter") paths$det_winter else paths$det_all

cameras_grid <- readr::read_csv(cams_file, show_col_types = FALSE) |>
  dplyr::select(Project_name, Station_ID, Latitude, Longitude,
                Start_Deployment_date, End_Deployment_date) |>
  tidyr::drop_na(Latitude, Longitude) |>
  make_sf_points(lon = "Longitude", lat = "Latitude",
                 crs_in = crs_wgs84, crs_out = crs_albers)

sp_cam <- readr::read_csv(det_file, show_col_types = FALSE) |>
  dplyr::select(Project_name, Species, Station_ID, Latitude, Longitude, Date_Time.y) |>
  tidyr::drop_na(Date_Time.y)
```

## Basemap & quick QA plots

```{r qa-plot}
# All points overlay for orientation
p_all <- ggplot() +
  geom_sf(data = columbian_area, fill = NA, color = "grey50") +
  geom_sf(data = dplyr::filter(subpopulations, Subpop != "Boreal"), aes(fill = Subpop), alpha = 0.3) +
  geom_sf(data = DNA, color = "firebrick", size = 1, alpha = 0.7) +
  geom_sf(data = cameras_grid, color = "purple", size = 1, alpha = 0.7) +
  ggtitle("DNA & Camera deployments over Columbian AOI") + theme_minimal()

p_all
```
# DNA sampling by project

```{r}
DNA_by_project <- ggplot()+
  geom_sf(data = columbian_area, fill = NA, color = "grey50") +
  geom_sf(data = dplyr::filter(subpopulations, Subpop != "Boreal"), aes(fill = Subpop), alpha = 0.3) +
  geom_sf(data = DNA,aes(color = Project_name), size = 1) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "bottom",       # move legend out of the way
    legend.key.width = unit(6, "pt"), # shrink keys
    legend.key.height = unit(6, "pt"),
    legend.text = element_text(size = 6),
    legend.title = element_text(size = 7)
  ) +
  guides(
    color = guide_legend(
      nrow = 3,                       # put legend items on one row
      byrow = TRUE,
      override.aes = list(size = 3)   # make legend points/lines smaller
    )
  )


DNA_by_project
```

# Camera sampling by project

```{r}
camera_by_project <- ggplot()+
  geom_sf(data = columbian_area, fill = NA, color = "grey50") +
  geom_sf(data = dplyr::filter(subpopulations, Subpop != "Boreal"), aes(fill = Subpop), alpha = 0.3) +
  geom_sf(data= cameras_grid,  aes(color = Project_name))+
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "bottom",       # move legend out of the way
    legend.key.width = unit(6, "pt"), # shrink keys
    legend.key.height = unit(6, "pt"),
    legend.text = element_text(size = 6),
    legend.title = element_text(size = 7)
  ) +
  guides(
    color = guide_legend(
      nrow = 3,                       # put legend items on one row
      byrow = TRUE,
      override.aes = list(size = 3)   # make legend points/lines smaller
    )
  )


camera_by_project
```
# Plot by project of interest

```{r}
project_to_explore <- "Chilcotin Cougar Project"

p_explore <- ggplot()+
  geom_sf(data = cameras_grid, color= "black", size = 1) +
  geom_sf(data= filter(cameras_grid, Project_name==project_to_explore), aes(color = Station_ID)) 
p_explore

```


## Project rules and state map
Based on earlier graphs, decide on state space assigment and change below accordingly
Two common renames are implemented programmatically. Update as needed.

```{r rules-states}

project_rules <- tibble::tibble(
  project_match = c(
    "Interior BC Moose Predator Camera Study",
    "Fisher Exclusion Box Program"
  ),
  rename_fun = list(
    function(df) dplyr::mutate(
      df,
      Project_name = ifelse(
        grepl("^PGS", Station_ID),
        "Interior BC Moose & Predator Camera Study-1",
        "Interior BC Moose & Predator Camera Study-2"
      )
    ),
    function(df) dplyr::mutate(
      df,
      Project_name = dplyr::case_when(
        stringr::str_detect(Station_ID, "^(WS|JD)") ~ "Fisher Exclusion Box Program-1",
        TRUE ~ "Fisher Exclusion Box Program-2"
      )
    )
  )
)


state_map <- tibble::tribble(
  ~Project_name,                                      ~State,
  "Grizzly Bear Corridor 2021",                       2,
  "Skeena Predator Project UBCO",                     3,
  "Fisher Exclusion Box Program-1",                   4,
  "Interior BC Moose & Predator Camera Study-1",      4,
  "Itcha Ilgachuz",                                   5,
  "South Chilcotin Mountains Camera Sampling",        5,
  "Chilcotin Cougar Project",                         5,
  "Chilcotin Cougar Project Prey Grid",               5,
  "Chilcotin Grizzy Bear Cameras 2021",               5,
  "Chilcotin Restoration Monitoring 2023",            5,
  "data_TNG trail cameras_2023.xlsx",                 5,
  "Esketemc Adaptive Management Project",             6,
  "BC SIM Deer",                                      6,
  "Interior BC Moose & Predator Camera Study-2",      6,
  "Fisher Exclusion Box Program-2",                   6
)
```


## Apply project rules & assign states
- For cameras and detections.
- Create a **short** camera set within AOI and de-duplicated to one row per station.

```{r apply-rules}
# Apply rules
cameras_grid <- apply_project_rules(cameras_grid, project_rules)
sp_cam       <- apply_project_rules(sp_cam, project_rules)

# AOI-filter & one row per Station_ID (deployment with earliest start)
cameras_grid_aoi <- sf::st_intersection(cameras_grid, columbian_area)

cameras_grid_short <- cameras_grid_aoi |>
  sf::st_drop_geometry() |>
  dplyr::arrange(Station_ID, Start_Deployment_date) |>
  dplyr::distinct(Station_ID, .keep_all = TRUE) |>
  make_sf_points(lon = "Longitude", lat = "Latitude", crs_in = crs_wgs84, crs_out = crs_albers)

# Optional programmatic outlier removal by ID/project (kept as per original intent)
cameras_to_remove <- c("witw14", "witw12", "Scat Study 05", "RC-06", "3087.1", "4210.5", "4210_R", "BCTA55")
cameras_grid_short1 <- cameras_grid_short |>
  dplyr::filter(Project_name != "Serb Creek West Babine Wildlife Cameras" | is.na(Project_name)) |>
  dplyr::filter(!Station_ID %in% cameras_to_remove)

# Assign states
cameras_grid_states <- assign_states(cameras_grid_short1, state_map)
DNA_states          <- assign_states(DNA, state_map)
sp_cam_states       <- assign_states(sp_cam, state_map)

# QA: any NA States?  ## should be 0, if not use code commented out below.
sum(is.na(cameras_grid_states$State))
sum(is.na(DNA_states$State))

cameras_na <- cameras_grid_states %>%
    dplyr::filter(is.na(State)) %>%
    dplyr::select(Project_name, Station_ID, Start_Deployment_date, End_Deployment_date) %>%
    arrange(Project_name, Station_ID)
unique(cameras_na$Project_name)
#Cap to first year per station (≤ 365 days)
deploy_df_cap <- cameras_grid_states %>%
  dplyr::mutate(
    Start_Deployment_date = as.Date(Start_Deployment_date),
    End_Deployment_date   = as.Date(End_Deployment_date),
    End_Deployment_date   = pmin(End_Deployment_date, Start_Deployment_date + 365 - 1)
  ) %>%
  dplyr::filter(End_Deployment_date >= Start_Deployment_date)

# ... then in the cam-ops-preserve chunk:
deploy_df <- deploy_df_cap   # <<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<< use the capped table

```

## Visual verification of state spaces

```{r verify-states}
pp <- ggplot() +
    geom_sf(data = columbian_area, fill = NA, color = "grey50") +
  geom_sf(data = dplyr::filter(subpopulations, Subpop != "Boreal"), aes(fill = Subpop), alpha = 0.5) +  
 # scale_fill_grey(start = 0.9, end = 0.4)+
    geom_sf(data = DNA_states, aes(color = as.factor(State.x)), size = 1) +
  geom_sf(data = cameras_grid_states, aes(color = as.factor(State)), size = 1, shape = 1) +
  ggtitle("DNA (points) and Cameras (rings) by State") +
  theme_minimal()
pp
```

## DNA occasions & `Y.s` array

#Split by state
```{r}
#create list of both databases based on state space
DNA_states$State <- DNA_states$State.x

cam_short_list <- split(cameras_grid_states, f= cameras_grid_states$State)

DNA_short_list <- split(DNA_states, f= DNA_states$State)
```

## Build DNA operability matrices O.s (stations x occasions, 0/1)
```{r}
## ---- Build DNA operability matrices O.s (stations x occasions, 0/1) ----
## Robust to chunk order:
##  - Computes 'occasion' locally (rleid(End_date))
##  - Recreates station ordering exactly as in Y.s (after species filter on Animal_ID)
##  - Returns a list O.s with one matrix per state; rows = stations, cols = occasions, values in {0,1}

# Helper: add occasions with the same rule used for Y.s
add_occasions <- function(dna_state_df) {
  if (!("End_date" %in% names(dna_state_df))) {
    stop("DNA table is missing 'End_date' needed to compute occasions.")
  }
  dna_state_df |>
    dplyr::group_by(State, Project_name, Station_ID) |>
    dplyr::arrange(State,
                   stringr::str_to_lower(Project_name),
                   Station_ID, End_date) |>
    dplyr::mutate(occasion = data.table::rleid(End_date)) |>
    dplyr::ungroup()
}

# Helper: station order identical to Y.s build
# (set Animal_ID = NA for non-fisher, then pivot wide and read the station columns)
get_station_order <- function(dna_state_df, species_of_interest) {
  nm <- dna_state_df |>
    dplyr::mutate(Animal_ID = dplyr::case_when(
      Species == species_of_interest ~ Animal_ID,
      TRUE ~ NA_character_
    )) |>
    sf::st_drop_geometry() |>
    dplyr::mutate(st_filler = "goop") |>
    dplyr::select(Station_ID, Animal_ID, st_filler) |>
    dplyr::distinct() |>
    tidyr::pivot_wider(
      names_from  = Station_ID,
      values_from = st_filler,
      names_prefix = "st_"
    ) |>
    dplyr::filter(!duplicated(Animal_ID)) |>
    dplyr::filter(!is.na(Animal_ID)) |>
    dplyr::arrange(Animal_ID) |>
    names()

  # If pivot produced only 'Animal_ID' (no station columns), return empty
  if (length(nm) <= 1L) return(character(0))
  nm <- nm[-1L]                     # drop 'Animal_ID'
  sub("^st_", "", nm)               # strip 'st_' to get raw Station_IDs
}

# Build one O.s matrix for a state's DNA table
build_Os_for_state <- function(dna_state_df, species_of_interest) {
  if (is.null(dna_state_df) || nrow(dna_state_df) == 0) {
    return(matrix(numeric(0), 0, 0))
  }

  # Compute local occasions (same rule as Y.s)
  dna_occ <- add_occasions(dna_state_df)

  # Number of occasions (K)
  K <- suppressWarnings(max(dna_occ$occasion, na.rm = TRUE))
  if (!is.finite(K) || K < 1L) {
    return(matrix(numeric(0), 0, 0))
  }

  # Station order aligned to Y.s (after species filter on Animal_ID)
  st_ids <- get_station_order(dna_occ, species_of_interest)
  if (length(st_ids) == 0L) {
    # No station columns present once aligned to Y.s
    return(matrix(numeric(0), 0, 0))
  }

  # Initialize matrix
  Os <- matrix(
    0L,
    nrow = length(st_ids),
    ncol = K,
    dimnames = list(st_ids, paste0("Occ_", seq_len(K)))
  )

  # A station is "operable = 1" in an occasion if it appears in the DNA table in that occasion
  # NOTE: Operability is not filtered by species — it reflects trap availability.
  for (k in seq_len(K)) {
    active <- dna_occ |>
      sf::st_drop_geometry() |>
      dplyr::filter(occasion == k) |>
      dplyr::distinct(Station_ID) |>
      dplyr::pull(Station_ID)
    if (length(active)) Os[st_ids %in% active, k] <- 1L
  }

  Os
}

# ---- Build O.s as a list (one matrix per state) ----
O.s <- vector("list", length(DNA_short_list))
names(O.s) <- names(DNA_short_list)

for (i in seq_along(DNA_short_list)) {
  O.s[[i]] <- build_Os_for_state(DNA_short_list[[i]], species_of_interest)
}

# ---- Optional: sanity checks vs Ys_list (if available) ----
# If you've run the Y.s chunk, check that O.s rows (= stations) and cols (= occasions) match
try({
  if (exists("Ys_list")) {
    for (i in seq_along(O.s)) {
      # Only check states that actually produced a Y.s (Ys_list entry not NULL)
      if (!is.null(Ys_list[[i]]) && length(Ys_list[[i]]) > 0) {
        # Some states may have 0 individuals but still a station set;
        # J and K from Ys_list are the intended station/occasion counts.
        if (nrow(O.s[[i]]) != Ys_list[[i]]$J || ncol(O.s[[i]]) != Ys_list[[i]]$K) {
          message(sprintf(
            "Note: state '%s' has O.s dim %s x %s, while Y.s expects %s x %s",
            names(O.s)[i], nrow(O.s[[i]]), ncol(O.s[[i]]),
            Ys_list[[i]]$J, Ys_list[[i]]$K
          ))
        }
      }
    }
  }
}, silent = TRUE)

# Optional quick summary to console (uncomment to inspect)
# print(lapply(O.s, dim))
```


```{r dna-occasions}
library(data.table)

# Build occasions (run-length ID by End_date) and construct Y.s per state
make_Ys_for_state <- function(dna_state_df) {
  dna_state_df <- dna_state_df |>
    dplyr::mutate(Animal_ID = dplyr::case_when(
      Species == species_of_interest ~ Animal_ID, TRUE ~ NA_character_
    )) |>
    dplyr::group_by(State, Project_name, Station_ID) |>
    dplyr::arrange(State, stringr::str_to_lower(Project_name), Station_ID, End_date) |>
    dplyr::mutate(occasion = data.table::rleid(End_date)) |>
    dplyr::ungroup()

  matrix_frame <- dna_state_df |>
    sf::st_drop_geometry() |>
    dplyr::mutate(st_filler = 'goop') |>
    dplyr::select(Station_ID, Animal_ID, st_filler) |>
    dplyr::distinct() |>
    tidyr::pivot_wider(names_from = Station_ID, values_from = st_filler, names_prefix = 'st_') |>
    dplyr::filter(!duplicated(Animal_ID)) |>
    dplyr::filter(!is.na(Animal_ID)) |>
    dplyr::arrange(Animal_ID)

  n_ind <- dplyr::n_distinct(stats::na.omit(matrix_frame$Animal_ID))
  K     <- max(dna_state_df$occasion)
  J     <- ncol(matrix_frame) - 1L

  make_occ_tbl <- function(k_) {
    dat <- dna_state_df |>
      dplyr::filter(occasion == k_) |>
      sf::st_drop_geometry() |>
      dplyr::filter(!is.na(Animal_ID)) |>
      dplyr::select(Animal_ID, Station_ID, Species) |>
      tidyr::pivot_wider(names_from = Station_ID, values_from = Species, names_prefix = 'st_') |>
      dplyr::mutate(dplyr::across(-Animal_ID, ~ dplyr::case_when(is.na(.x) ~ 0L, .x == species_of_interest ~ 1L, TRUE ~ 0L)))

    st_cols <- names(dat)[-1]
    filler  <- matrix_frame |>
      dplyr::select(-all_of(st_cols)) |>
      dplyr::mutate(dplyr::across(-Animal_ID, ~ 0L))

    out <- dplyr::left_join(dat, filler, by = "Animal_ID") |>
      dplyr::bind_rows(filler[!filler$Animal_ID %in% dat$Animal_ID, ]) |>
      dplyr::mutate(dplyr::across(-Animal_ID, ~ ifelse(is.na(.x), 0L, .x))) |>
      dplyr::arrange(Animal_ID) |>
      dplyr::select(Animal_ID, names(matrix_frame)[-1])
    out
  }

  occ_tbls <- purrr::map(1:K, make_occ_tbl)
  Y_s <- array(0L, dim = c(n_ind, J, K))
  for (k in seq_len(K)) Y_s[,,k] <- as.matrix(occ_tbls[[k]][,-1L])
  list(Y_s = Y_s, n_ind = n_ind, J = J, K = K)
}

Ys_list <- purrr::map(DNA_short_list, make_Ys_for_state)
Y.s <- purrr::map(Ys_list, "Y_s")
```

## Camera operation & detection histories — **preserve zero-detection projects**
This section ensures that **every project with deployments** contributes to `detectionHistory()` even if it has **no detections**. We achieve this by adding a **single placeholder record** to those projects before running `detectionHistory()`, and then **zero out** the resulting detections so that counts remain correct but matrix shapes include all projects.

```{r cam-ops-preserve}

# ---- Camera operation & detection histories (preserve zero-detection projects) ----
# This chunk yields an outer list DetHist of length = number of states with deployments (often 4)
# and inner list lengths equal to the number of projects per state, e.g. c(1, 2, 6, 4).

# Use the same deployment/detection objects as your pipeline
deploy_df <- if (exists("cameras_grid_short_states")) cameras_grid_short_states else cameras_grid_states
det_df    <- sp_cam_states

# Safety default: preserve projects that have deployments but zero detections
if (!exists("preserve_zero_projects", inherits = FALSE)) preserve_zero_projects <- TRUE

# camtrapR wants the datetime column as character
det_df$Species    <- species_of_interest
det_tbl           <- as.data.frame(det_df)
det_tbl$Date_Time2 <- as.character(det_tbl$Date_Time.y)

# Split by STATE (only states that actually have deployments remain, e.g., 4 states)
camgrid_by_state <- split(deploy_df, f = deploy_df$State)
cam_by_state     <- split(det_tbl,  f = det_tbl$State)

CampOp.list <- vector("list", length(camgrid_by_state))
DetHist     <- vector("list", length(camgrid_by_state))
names(CampOp.list) <- names(camgrid_by_state)
names(DetHist)     <- names(camgrid_by_state)

for (s in seq_along(camgrid_by_state)) {

  # -------------------- Deployments by project (within state s) --------------------
  cam_df <- camgrid_by_state[[s]] %>%
    dplyr::filter(!is.na(Project_name)) %>%
    dplyr::arrange(stringr::str_to_lower(Project_name), Station_ID)

  # Split using the SAME arranged data frame (prevents misalignment)
  cam_deploy_by_proj <- split(cam_df, f = cam_df$Project_name, drop = TRUE)
  all_projects       <- names(cam_deploy_by_proj)

  # cameraOperation per project
  CampOp.list[[s]] <- lapply(
    cam_deploy_by_proj,
    function(df_proj) {
      camtrapR::cameraOperation(
        CTtable        = sf::st_drop_geometry(df_proj),
        stationCol     = "Station_ID",
        setupCol       = "Start_Deployment_date",
        retrievalCol   = "End_Deployment_date",
        hasProblems    = FALSE,
        dateFormat     = "ymd",
        occasionStartTime = 0
      )
    }
  )

  # -------------------- Detections by project (within state s) --------------------
  if (length(cam_by_state) >= s && !is.null(cam_by_state[[s]]) && nrow(cam_by_state[[s]]) > 0) {
    det_state <- cam_by_state[[s]] %>%
      dplyr::filter(!is.na(Project_name)) %>%
      dplyr::arrange(stringr::str_to_lower(Project_name), Station_ID)
    det_by_proj <- split(det_state, f = det_state$Project_name, drop = TRUE)
  } else {
    det_by_proj <- list()
  }

  # Which projects have deployments but no detections?
  det_projects     <- names(det_by_proj)
  missing_projects <- setdiff(all_projects, det_projects)

  # Filter detection rows to stations present in camOp rownames (required by detectionHistory)
  for (p in det_projects) {
    camop <- CampOp.list[[s]][[p]]
    if (is.null(camop)) next
    allowed <- rownames(camop)
    det_by_proj[[p]] <- det_by_proj[[p]] %>%
      dplyr::filter(Station_ID %in% allowed)
    # If all detections are filtered out, treat project as missing
    if (nrow(det_by_proj[[p]]) == 0) {
      missing_projects <- union(missing_projects, p)
      det_by_proj[[p]] <- NULL
    }
  }

  # Add exactly one placeholder (anchored at a valid deployment station) for zero-detection projects
  if (preserve_zero_projects && length(missing_projects)) {
    for (p in missing_projects) {
      camop <- CampOp.list[[s]][[p]]
      if (is.null(camop)) next
      allowed <- rownames(camop)
      if (length(allowed) == 0) next
      # Use the first station in camOp for this project
      stn <- allowed[1]
      # Find a reasonable date from deployments (start date for that station); fallback = today
      dep_proj_df <- sf::st_drop_geometry(cam_deploy_by_proj[[p]])
      dep_stn     <- dplyr::filter(dep_proj_df, Station_ID == stn)
      dt0         <- if (nrow(dep_stn) > 0) as.Date(dep_stn$Start_Deployment_date[1]) else Sys.Date()

      det_by_proj[[p]] <- data.frame(
        Project_name = p,
        Species      = species_of_interest,
        Station_ID   = stn,
        Date_Time2   = paste0(as.character(dt0), " 01:42:00"),
        stringsAsFactors = FALSE
      )
    }
  }

  # -------------------- Build detection histories per project --------------------
  DetHist_state <- vector("list", length(all_projects))
  names(DetHist_state) <- all_projects

  for (p in all_projects) {
    camop <- CampOp.list[[s]][[p]]
    if (is.null(camop)) next
    rt <- det_by_proj[[p]]
    if (is.null(rt) || nrow(rt) == 0) next

    # Final guard: keep only stations present in camOp
    rt <- rt[rt$Station_ID %in% rownames(camop), , drop = FALSE]
    if (nrow(rt) == 0) next

    DetHist_state[[p]] <- camtrapR::detectionHistory(
      recordTable          = rt,
      camOp                = camop,
      stationCol           = "Station_ID",
      speciesCol          = "Species",
      recordDateTimeCol    = "Date_Time2",
      species              = species_of_interest,
      occasionLength       = occasion_length,
      day1                 = "station",
      datesAsOccasionNames = FALSE,
      includeEffort        = FALSE,
      timeZone             = tz_local
    )
  }

  # Zero-out placeholders to preserve all-zero projects (matrix exists, but counts remain correct)
  if (preserve_zero_projects && length(missing_projects)) {
    for (p in missing_projects) {
      if (!is.null(DetHist_state[[p]])) {
        dh <- DetHist_state[[p]]$detection_history
        dh[dh == 1] <- 0
        DetHist_state[[p]]$detection_history <- dh
      }
    }
  }

  # Store per-project results for this state
  DetHist[[s]] <- DetHist_state
}

# ---- Optional: quick structure check (uncomment to verify) ----
# lengths_per_state <- vapply(DetHist, function(x) length(Filter(Negate(is.null), x)), integer(1))
# print(lengths_per_state)   # expected something like c(1, 2, 6, 4)
```

## 13) Build camera matrices `O.o` (detections) and `O.o2` (effort)
```{r build-oo}

#Build O.o (detections: 0/1/NA) and O.o2 (effort: 1/0) per state
O.o  <- vector("list", length(DetHist)); names(O.o)  <- names(DetHist)
O.o2 <- vector("list", length(DetHist)); names(O.o2) <- names(DetHist)
df.list <- vector("list", length(DetHist))

for (s in seq_along(DetHist)) {
  # If no projects in this state, store empty 0x0 matrices and continue
  if (length(DetHist[[s]]) == 0) {
    O.o[[s]]  <- matrix(numeric(0), 0, 0)
    O.o2[[s]] <- matrix(numeric(0), 0, 0)
    next
  }

  # Row-bind project detection_history matrices; ldply adds an '.id' column by default
  df.list[[s]] <- plyr::ldply(
    DetHist[[s]],
    function(x) as.data.frame(x$detection_history),
    .id = ".id"
  )

  # Drop the id column (robustly)
  dh_df <- df.list[[s]]
  if (".id" %in% names(dh_df)) {
    dh_df <- dh_df[, setdiff(names(dh_df), ".id"), drop = FALSE]
  } else if (!is.numeric(dh_df[[1]])) {
    # Fallback: if first column isn't numeric, drop it
    dh_df <- dh_df[, -1, drop = FALSE]
  }

  # Coerce to numeric matrix
  mat <- suppressWarnings(as.matrix(dh_df))
  storage.mode(mat) <- "numeric"   # non-numeric cells become NA

  # Restrict to {0, 1, NA}: any non-zero (and non-NA) becomes 1; zeros stay 0; NAs stay NA
  mat[!is.na(mat) & mat != 0] <- 1

  # Save detections matrix (0/1/NA)
  O.o[[s]] <- mat

  # Effort matrix (1 where not NA, else 0) — optional but included for completeness
  eff <- mat
  eff[!is.na(eff)] <- 1
  eff[is.na(eff)]  <- 0
  O.o2[[s]] <- eff
}

# Optional sanity checks
# str(O.o[[1]]); str(O.o2[[1]])
# table(O.o[[1]], useNA = "ifany")

```


```{r}
## ---- Build Y.o (camera detections by state; alias of O.o) ----
## Goal: 6-slot list named "1".."6" so downstream code always has all state slots.
## Semantics: Y.o[[s]] = O.o[[s]] when cameras exist in state s; else NULL.

state_ids <- as.character(1:6)

# Ensure O.o is keyed by state id strings; we built it using names(DetHist),
# which are the states that actually have deployments (e.g., "3","4","5","6").
# We'll expand that into a full 6-slot Y.o.
Y.o <- setNames(vector("list", length(state_ids)), state_ids)

for (s in state_ids) {
  if (!is.null(O.o[[s]]) && length(O.o[[s]]) > 0) {
    Y.o[[s]] <- O.o[[s]]
  } else {
    # keep NULL (as your original script did for states without cameras)
    Y.o[[s]] <- NULL
  }
}

# (Optional) If you prefer empty 0x0 matrices instead of NULLs, replace the "else" with:
# Y.o[[s]] <- matrix(numeric(0), 0, 0)
```

## Coordinates for SCR (km -> m)
```{r scr-inputs}
# DNA traps (X1) and Cameras (X2)
Traps_list <- purrr::imap(DNA_short_list, ~ .x |>
  dplyr::select(Project_name, Station_ID, UTM_X, UTM_Y) |>
  sf::st_drop_geometry() |>
  dplyr::arrange(stringr::str_to_lower(Project_name), Station_ID) |>
  dplyr::distinct() |>
  dplyr::select(UTM_X, UTM_Y) |>
  as.matrix() / coord_scale_km)

cam_list <- purrr::imap(cam_short_list, ~ .x |>
  dplyr::select(Station_ID, UTM_X, UTM_Y) |>
  sf::st_drop_geometry() |>
  dplyr::distinct() |>
  dplyr::select(UTM_X, UTM_Y) |>
  as.matrix() / coord_scale_km)

# Expect exactly 6 states named "1","2","3","4","5","6"
n_state_space <- 6
state_ids <- as.character(1:n_state_space)

# empty 0x2 matrix placeholder (km)
make_empty_xy_km <- function() {
  matrix(numeric(0), nrow = 0, ncol = 2,
         dimnames = list(NULL, c("UTM_X","UTM_Y")))
}

# Ensure Traps_list has entries for all 6 states, in order
Traps_list <- setNames(Traps_list[state_ids], state_ids)
Traps_list <- purrr::imap(Traps_list, ~ if (is.null(.x)) make_empty_xy_km() else .x)

# Ensure cam_list has entries for all 6 states, in order
cam_list <- setNames(cam_list[state_ids], state_ids)
cam_list <- purrr::imap(cam_list, ~ if (is.null(.x)) make_empty_xy_km() else .x)

## ---- Bounds per state (simple, programmatic; 50 km padding; km units) ----

# Assumes:
# - state_ids <- as.character(1:n_state_space) already defined
# - Traps_list and cam_list are lists of length 6 (states "1".."6"), each entry is a
#   0x2 or Nx2 matrix in KM with columns UTM_X, UTM_Y (as in your current setup)

# Helper: combine two XY matrices; return 0x2 if both empty
combine_xy <- function(a, b) {
  # Coerce NULLs to 0x2 matrices
  to_xy <- function(m) {
    if (is.null(m) || length(m) == 0) {
      out <- matrix(numeric(0), nrow = 0, ncol = 2)
      colnames(out) <- c("UTM_X","UTM_Y")
      return(out)
    }
    # ensure numeric 2 cols with names
    m <- as.matrix(m[, 1:2, drop = FALSE])
    storage.mode(m) <- "numeric"
    if (is.null(colnames(m))) colnames(m) <- c("UTM_X","UTM_Y")
    m
  }
  a <- to_xy(a); b <- to_xy(b)
  if (nrow(a) == 0 && nrow(b) == 0) return(a)
  if (nrow(a) == 0) return(b)
  if (nrow(b) == 0) return(a)
  rbind(a, b)
}

# Compute bounds with +/- 50 km padding
compute_bounds_km <- function(m, pad_km = 50) {
  if (is.null(m) || nrow(m) == 0) {
    return(list(xlim = c(NA_real_, NA_real_), ylim = c(NA_real_, NA_real_), area = NA_real_))
  }
  x <- m[, 1]; y <- m[, 2]
  xmin <- min(x, na.rm = TRUE) - pad_km
  xmax <- max(x, na.rm = TRUE) + pad_km
  ymin <- min(y, na.rm = TRUE) - pad_km
  ymax <- max(y, na.rm = TRUE) + pad_km
  list(
    xlim = c(xmin, xmax),
    ylim = c(ymin, ymax),
    area = (xmax - xmin) * (ymax - ymin)
  )
}

# Build per-state combined XY (traps ∪ cameras within the same state), then bounds
xlim.list <- setNames(vector("list", length(state_ids)), state_ids)
ylim.list <- setNames(vector("list", length(state_ids)), state_ids)
Area.list <- setNames(vector("list", length(state_ids)), state_ids)

for (s in state_ids) {
  xy_s <- combine_xy(Traps_list[[s]], cam_list[[s]])   # per-state union (no cross-state assumptions)
  b    <- compute_bounds_km(xy_s, pad_km = 50)
  xlim.list[[s]] <- b$xlim
  ylim.list[[s]] <- b$ylim
  Area.list[[s]] <- b$area
}

# Convenience 6x2 matrices (km) and a named area vector (km^2)
xlims <- do.call(rbind, xlim.list); rownames(xlims) <- state_ids
ylims <- do.call(rbind, ylim.list); rownames(ylims) <- state_ids
Areas <- setNames(
  vapply(Area.list, function(z) if (length(z)) z else NA_real_, numeric(1)),
  state_ids
)

# Optional checks
# print(xlims); print(ylims); print(Areas)

# Build X1 (DNA traps) and X2 (cameras) in meters (km -> m), preserving 6 slots
X1 <- purrr::map(Traps_list, ~ .x * 1000)
X2 <- purrr::map(cam_list,   ~ .x * 1000)

# Sanity checks
stopifnot(length(X1) == 6, length(X2) == 6)
stopifnot(identical(names(X1), state_ids), identical(names(X2), state_ids))
```

#Final map

```{r}
# --- INPUT ASSUMPTIONS (already loaded in your session) ---
# columbian_area          : sf POLYGON(s) for the broader region
# subpopulations          : sf POLYGON(s) with column 'Subpop'
# DNA_states              : sf POINT(s) with column 'State.x' (DNA sites)
# cameras_grid_states     : sf POINT(s) with column 'State'   (camera sites)
# state_spaces (optional) : sf POLYGON(s) with 'session' or 'State' per state space
# xlim, ylim (optional)   : 6x2 matrices with session extents; may be meters or kilometers

# --- 1) Transform ALL layers to a projected CRS (meters) for clean zooming ---
# (BC Albers is EPSG:3005; adjust if you use a different project CRS)
target_crs <- 3005
columbian_area       <- st_transform(columbian_area, target_crs)
subpopulations       <- st_transform(subpopulations, target_crs)
DNA_states           <- st_transform(DNA_states, target_crs)
cameras_grid_states  <- st_transform(cameras_grid_states, target_crs)
if (exists("state_spaces")) {
  state_spaces <- st_transform(state_spaces, target_crs)
}

# --- 2) Harmonize and lock the state factor levels across layers ---
# Rename DNA 'State.x' -> 'State' so both use the same aesthetic.
DNA_states2 <- DNA_states %>% dplyr::select(-State) %>%  dplyr::rename(State = State.x)
# 

#optiona--(save to shapefile for ESRI map)--
#st_write(DNA_states2, "DNA_states.gpkg", layer = "points", delete_layer = TRUE)
st_write(cameras_grid_states, "cameras_grid_states.gpkg", layer = "points", delete_layer = TRUE)

# Build a shared, explicit set of state levels from BOTH layers (order fixed).
state_levels <- sort(unique(c(
  as.character(DNA_states2$State),
  as.character(cameras_grid_states$State)
)))

# Relevel both layers to use the same factor levels (prevents palette mismatches).
DNA_states2$State          <- factor(as.character(DNA_states2$State), levels = state_levels)
cameras_grid_states$State  <- factor(as.character(cameras_grid_states$State), levels = state_levels)

# --- 3) Define a single shared color palette for states & apply it everywhere ---
# You can swap to viridis or ColorBrewer if you prefer.
pal <- scales::hue_pal()(length(state_levels))
names(pal) <- state_levels

shared_color_scale <- scale_color_manual(
  name   = "State",
  values = pal,
  breaks = state_levels,
  limits = state_levels,
  drop   = FALSE
)

# --- 4) Build the main map (pp) using the shared color scale ---
pp <- ggplot() +
  geom_sf(data = columbian_area, fill = NA, color = "grey50") +
  geom_sf(
    data = dplyr::filter(subpopulations, Subpop != "Boreal"),
    aes(fill = Subpop), alpha = 0.5
  ) +
  geom_sf(
    data = DNA_states2,
    aes(color = State),
    size = 1.5, shape= 1
  ) +
  geom_sf(
    data = cameras_grid_states,
    aes(color = State),
    size = 1, shape = 3, stroke = 1.1   # rings for cameras
  ) +
  shared_color_scale +                     # << enforce the exact same colors
  ggtitle("DNA (points) and Cameras (cross) by State Space") +
  theme_minimal(base_size = 11) +
  guides(
    color = guide_legend(title = "State"),
    fill  = guide_legend(title = "Subpopulation"))+
    theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
    panel.background = element_blank()
  )

# --- 5) Helper: Zoom panel function (same palette + robust extents + sf-native jitter) ---
make_zoom_plot <- function(
  g,
  xlim_mat = NULL, ylim_mat = NULL,        # optional matrices (6x2)
  extents_in_km = FALSE,                   # set TRUE if your xlim/ylim are in kilometers
  dna_sf = DNA_states2,                    # sf POINTS with 'State'
  cams_sf = cameras_grid_states,           # sf POINTS with 'State'
  state_poly = NULL,                       # optional sf POLYGON(s) with 'session' or 'State'
  title_prefix = "State space",
  jitter_amount = 10,                      # meters; requires projected CRS
  min_pad = 400                            # minimum panel width/height in meters
) {
  # Subset per state/session (g)
  dna_g  <- dna_sf  |> dplyr::filter(State == state_levels[g])
  cams_g <- cams_sf |> dplyr::filter(State == state_levels[g])

  # Optional: corresponding state polygon/boundary
  poly_g <- NULL
  if (!is.null(state_poly) && inherits(state_poly, "sf")) {
    if ("session" %in% names(state_poly)) {
      # If sessions map 1..6 to the same order as state_levels, you can match by index here.
      poly_g <- state_poly |> dplyr::filter(session == g)
    } else if ("State" %in% names(state_poly)) {
      poly_g <- state_poly |> dplyr::filter(State == state_levels[g])
    }
  }

  # Light jitter to separate coincident points (in meters)
  if (nrow(dna_g) > 0 && jitter_amount > 0)  dna_g  <- sf::st_jitter(dna_g,  amount = jitter_amount)
  if (nrow(cams_g) > 0 && jitter_amount > 0) cams_g <- sf::st_jitter(cams_g, amount = jitter_amount)

  # Decide zoom extents:
  use_mats <- !is.null(xlim_mat) && !is.null(ylim_mat) &&
              (is.matrix(xlim_mat) || is.data.frame(xlim_mat)) &&
              (is.matrix(ylim_mat) || is.data.frame(ylim_mat)) &&
              nrow(as.data.frame(xlim_mat)) >= g && nrow(as.data.frame(ylim_mat)) >= g

  if (use_mats) {
    xlim_use <- as.numeric(xlim_mat[g, ])
    ylim_use <- as.numeric(ylim_mat[g, ])
    if (isTRUE(extents_in_km)) {
      xlim_use <- xlim_use * 1000
      ylim_use <- ylim_use * 1000
    }
  } else {
    # Fallback: bbox from available layers for this state
    candidates <- list(dna_g, cams_g, poly_g)
    bxs <- lapply(candidates, function(x) if (inherits(x, "sf") && nrow(x) > 0) sf::st_bbox(x) else NULL)
    bxs <- Filter(Negate(is.null), bxs)
    if (length(bxs) == 0) stop(sprintf("No data to determine bbox for state/session %d", g))
    xlim_use <- range(sapply(bxs, function(b) c(b["xmin"], b["xmax"])))
    ylim_use <- range(sapply(bxs, function(b) c(b["ymin"], b["ymax"])))
  }

  # Ensure spans aren’t degenerate; pad tiny ranges
  pad_if_tiny <- function(lims, min_pad) {
    span <- diff(lims)
    if (!is.finite(span) || span < min_pad) {
      mid <- mean(lims)
      return(c(mid - min_pad/2, mid + min_pad/2))
    }
    pad <- span * 0.05
    c(lims[1] - pad, lims[2] + pad)
  }
  xlim_use <- pad_if_tiny(xlim_use, min_pad = min_pad)
  ylim_use <- pad_if_tiny(ylim_use, min_pad = min_pad)

  # Build the zoom panel with the SAME color scale
  pz <- ggplot()
  if (!is.null(poly_g) && nrow(poly_g) > 0) {
    pz <- pz + geom_sf(data = poly_g, fill = NA, color = "black", linewidth = 0.5)
  }
  if (nrow(dna_g) > 0) {
    pz <- pz + geom_sf(data = dna_g, aes(color = State), shape= 1, size = 1)
  }
  if (nrow(cams_g) > 0) {
    pz <- pz + geom_sf(data = cams_g, color= 'black', size = 0.5, shape = 3, stroke = 1)
  }

 # show longitude/latitude ticks in panes (axes in degrees) ---

# 1) Build a rectangle (in your projected CRS) from the current x/y limits:
rect_proj <- sf::st_as_sfc(sf::st_bbox(
  c(xmin = xlim_use[1], xmax = xlim_use[2],
    ymin = ylim_use[1], ymax = ylim_use[2]),
  crs = sf::st_crs(target_crs)   # <- your projected CRS (e.g., 3005)
))

# 2) Transform that rectangle to geographic (EPSG:4326):
rect_geo <- sf::st_transform(rect_proj, 4326)
bb_geo   <- sf::st_bbox(rect_geo)  # now in lon/lat

# 3) Plot with geographic axes but keep geometries as-is (projected):
pz +
  #shared_color_scale +
  coord_sf(
    crs  = sf::st_crs(4326),                        # axes in lon/lat
    xlim = c(bb_geo["xmin"], bb_geo["xmax"]),       # degrees
    ylim = c(bb_geo["ymin"], bb_geo["ymax"]),       # degrees
    expand = FALSE,
    label_graticule = "EWNS"                         # optional: label all sides
  ) +
  # 4) (Optional) fewer, cleaner longitude labels:
  scale_x_continuous(breaks = scales::breaks_extended(n = 3),
                     labels = function(x) paste0(round(x, 3), "°")) +
  scale_y_continuous(breaks = scales::breaks_extended(n = 3),
                     labels = function(x) paste0(round(x, 3), "°")) +
  ggtitle(sprintf("%s %s", title_prefix, state_levels[g])) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position   = "none",
    plot.title        = element_text(face = "bold", hjust = 0.5),
    panel.grid.major  = element_line(linewidth = 0.1, color = "grey85"),
    axis.text         = element_text(size = 8),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
    panel.background = element_blank()
  )
  
}
pp
```

```{r}
# --- Number of panes: use 6 or the number of distinct states you want to show ---
N.session <- min(6, length(state_levels))

# OPTIONAL: If you have xlim/ylim from SCR prep in the SAME CRS, pass them here.
# If those extents are in kilometers, set extents_in_km = TRUE below.
xlim_mat <- if (exists("xlim") && (is.matrix(xlim) || is.data.frame(xlim))) xlim else NULL
ylim_mat <- if (exists("ylim") && (is.matrix(ylim) || is.data.frame(ylim))) ylim else NULL

zoom_list <- lapply(seq_len(N.session), function(g) {
  make_zoom_plot(
    g = g,
    xlim_mat = xlim_mat,
    ylim_mat = ylim_mat,
    extents_in_km = TRUE,     # <-- set TRUE if your xlim/ylim are in kilometers
    dna_sf = DNA_states2,
    cams_sf = cameras_grid_states,
    state_poly = if (exists("state_spaces")) state_spaces else NULL,
    title_prefix = "State space",
    jitter_amount = 10,        # increase to 20–30 if points still overlap
    min_pad = 400              # broaden tiny extents to at least 400 m
  )
})

# --- Arrange six panes (2 columns), then combine with the main map ---
right_panel <- cowplot::plot_grid(
  plotlist = zoom_list,
  ncol = 2,
  align = "hv"
)

combined <- cowplot::plot_grid(
  pp, right_panel,
  ncol = 2,
  rel_widths = c(1.3, 1)   # make main map a bit wider than the panes
)

# Show it
print(combined)

# Save it
ggsave("DNA_Cameras_with_6_zoom_panes_matched_colors.png",
       combined, width = 14, height = 8, dpi = 300)
```


## Save workspace
```{r save}
if (!dir.exists(dirname(paste0(".",paths$save_rdata_winter)))) dir.create(dirname(paste0(".",paths$save_rdata_winter)), recursive = TRUE, showWarnings = FALSE)
if (!dir.exists(dirname(paste0(".",paths$save_rdata_all)))) dir.create(dirname(paste0(".",paths$save_rdata_all)), recursive = TRUE, showWarnings = FALSE)

if (identical(season, "winter")) {
  save.image(file = paste0(".",paths$save_rdata_winter))
  cat("Saved:", paste0(".",paths$save_rdata_winter), "\n")
} else {
  save.image(file = paste0(".",paths$save_rdata_all))
  cat("Saved:", paste0(".",paths$save_rdata_all, "\n"))
}
```


## Summary stations
```{r}
## ---- summary-stations, message=FALSE, warning=FALSE ---------------------

# Optional: set to TRUE to limit counts to stations within the Columbian AOI.
count_within_aoi <- FALSE

# Helper to intersect with AOI if requested and available
maybe_intersect_aoi <- function(sf_obj, do_intersect = FALSE, aoi = NULL) {
  if (!do_intersect) return(sf_obj)
  if (!inherits(sf_obj, "sf") || is.null(aoi) || !inherits(aoi, "sf")) return(sf_obj)
  tryCatch(sf::st_intersection(sf_obj, aoi), error = function(e) sf_obj)
}

# 1) Prepare unique station tables (1 row per Station_ID)
dna_sf_use <- maybe_intersect_aoi(DNA_states, do_intersect = count_within_aoi, aoi = columbian_area)
cam_sf_use <- cameras_grid_states  # already AOI-filtered upstream

dna_sites <- dna_sf_use |>
  sf::st_drop_geometry() |>
  dplyr::filter(!is.na(Station_ID)) |>
  dplyr::distinct(Station_ID, .keep_all = TRUE)

cam_sites <- cam_sf_use |>
  sf::st_drop_geometry() |>
  dplyr::filter(!is.na(Station_ID)) |>
  dplyr::distinct(Station_ID, .keep_all = TRUE)

# 2) Totals
n_dna_total <- nrow(dna_sites)
n_cam_total <- nrow(cam_sites)
n_dna_total
n_cam_total

# Helper: numeric key for sorting states like "State 1", "S-2", etc.
state_key <- function(x) suppressWarnings(as.integer(readr::parse_number(as.character(x))))

# 3) By State
dna_by_state <- dna_sites |>
  dplyr::mutate(State = dplyr::coalesce(as.character(State.x), as.character(State))) |>
  dplyr::mutate(State = ifelse(is.na(State) | State == "", "Unassigned", State)) |>
  dplyr::count(State, name = "n_DNA_stations") |>
  dplyr::arrange(state_key(State))

cam_by_state <- cam_sites |>
  dplyr::mutate(State = ifelse(is.na(State) | State == "", "Unassigned", as.character(State))) |>
  dplyr::count(State, name = "n_camera_traps") |>
  dplyr::arrange(state_key(State))

by_state <- dplyr::full_join(dna_by_state, cam_by_state, by = "State") |>
  dplyr::mutate(
    n_DNA_stations  = dplyr::coalesce(n_DNA_stations, 0L),
    n_camera_traps  = dplyr::coalesce(n_camera_traps, 0L)
  ) |>
  dplyr::arrange(state_key(State))
by_state

# 4) By Project (unchanged)
dna_by_project <- dna_sites |>
  dplyr::mutate(Project_name = ifelse(is.na(Project_name) | Project_name == "", "Unassigned", Project_name)) |>
  dplyr::count(Project_name, name = "n_DNA_stations") |>
  dplyr::arrange(dplyr::desc(n_DNA_stations), Project_name)
dna_by_project
```

## Summary 2
```{r}
## ---- camera_active_window_and_summary, message=FALSE, warning=FALSE ----

# Pick the camera table you already created (after filtering/exclusions)
cam_df <- cameras_grid_short1   # or: cam_df <- cameras_grid_states

# Ensure dates are Date, and drop rows without both dates
cam_df <- cam_df |>
  dplyr::mutate(
    Start_Deployment_date = as.Date(Start_Deployment_date),
    End_Deployment_date   = as.Date(End_Deployment_date)
  ) |>
  dplyr::filter(!is.na(Start_Deployment_date), !is.na(End_Deployment_date))

# Guard: if nothing remains, stop early with a helpful message
if (nrow(cam_df) == 0) stop("No camera deployments with valid start/end dates found in cam_df.")

# ---- Global active window (for narrative text) ----
active_start <- min(cam_df$Start_Deployment_date, na.rm = TRUE)
active_end   <- max(cam_df$End_Deployment_date,   na.rm = TRUE)

# Nicely formatted for manuscript text (Month YYYY)
active_start_fmt <- format(active_start, "%B %Y")
active_end_fmt   <- format(active_end,   "%B %Y")

# ---- Effort per station (camera-days) and summary stats ----
cam_df <- cam_df |>
  dplyr::mutate(camera_days = as.integer(End_Deployment_date - Start_Deployment_date + 1L)) |>
  dplyr::filter(camera_days > 0)

j_traps <- nrow(cam_df)

eff_min    <- min(cam_df$camera_days, na.rm = TRUE)
eff_max    <- max(cam_df$camera_days, na.rm = TRUE)
eff_mean   <- round(mean(cam_df$camera_days, na.rm = TRUE), 1)
eff_median <- median(cam_df$camera_days, na.rm = TRUE)
eff_q1     <- as.numeric(quantile(cam_df$camera_days, 0.25, na.rm = TRUE))
eff_q3     <- as.numeric(quantile(cam_df$camera_days, 0.75, na.rm = TRUE))

# ------------------------------------------------------------------
# k from analysis matrices, not global span
# ------------------------------------------------------------------

# Report k directly from the analysis matrices
K_state <- vapply(O.o, function(m) if (length(m)) ncol(m) else 0L, integer(1))
k_biweekly <- max(K_state, na.rm = TRUE)

# Optionally, also compute effort stats from the capped deployments
eff_cap_days <- deploy_df_cap %>%
  dplyr::mutate(camera_days = as.integer(End_Deployment_date - Start_Deployment_date + 1L)) %>%
  dplyr::filter(camera_days > 0)

j_traps  <- nrow(eff_cap_days)
eff_min  <- min(eff_cap_days$camera_days, na.rm = TRUE)
eff_max  <- max(eff_cap_days$camera_days, na.rm = TRUE)
eff_mean <- round(mean(eff_cap_days$camera_days, na.rm = TRUE), 1)

active_start_fmt <- format(min(eff_cap_days$Start_Deployment_date, na.rm = TRUE), "%B %Y")
active_end_fmt   <- format(max(eff_cap_days$End_Deployment_date,   na.rm = TRUE), "%B %Y")

# ---- Output: a one-line sentence ready to paste into Methods ----
cat("\n### Camera survey window & summary\n\n")
cat(sprintf(
  "Camera surveys were active from %s to %s, with effort ranging from %s to %s camera-days (mean = %s camera-days). ",
  active_start_fmt,
  active_end_fmt,
  format(eff_min, big.mark = ","),
  format(eff_max, big.mark = ","),
  format(eff_mean, big.mark = ",")
))
cat(sprintf(
  "We aggregated detections bi-weekly to construct an occupancy matrix (j = %s traps; k = %s bi-weekly occasions).",
  format(j_traps, big.mark = ","),
  k_biweekly
))
cat("\n")

# (Optional) return a list invisibly if you want to use values programmatically later
invisible(list(
  active_start = active_start_fmt,
  active_end   = active_end_fmt,
  j_traps      = j_traps,
  k_biweekly   = k_biweekly,
  eff_stats    = c(
    min = eff_min, q1 = eff_q1, median = eff_median,
    mean = eff_mean, q3 = eff_q3, max = eff_max
  )
))
```

#Table 1 
```{r}
## ==== Camera & DNA project × state summaries (consistent with Y.o) ====

# Packages
library(dplyr)
library(tidyr)
library(lubridate)
library(stringr)
library(purrr)
library(tibble)
library(knitr)

# Small helper for NULL-coalescing (used for DNA dates)
`%||%` <- function(a, b) if (!is.null(a)) a else b

# ---- 0) Canonical mapping: State -> SS[x] label + Subpopulation ----
canon_map <- tibble::tibble(
  State = 1:6,
  StateSpace = paste0("SS[", State, "]"),
  Subpopulation = c("Omineca","Omineca","Omineca","Omineca","Chilcotin","Cariboo")
)

# Helper: format "Year used"
fmt_year_used <- function(start_vec, end_vec) {
  s <- suppressWarnings(lubridate::as_date(start_vec))
  e <- suppressWarnings(lubridate::as_date(end_vec))
  y1 <- year(s); y2 <- year(e)
  out <- ifelse(is.finite(y1) & is.finite(y2),
                ifelse(y1 == y2, as.character(y1), paste0(y1, "–", y2)),
                "unknown")
  out
}

# =======================================================================================
# 1) CAMERA TABLE — project × state, with TotalDetections pulled from DetHist (→ Y.o)
# =======================================================================================

# Use the capped deployments if available (as in your pipeline)
cam_deploy_use <- if (exists("deploy_df_cap")) deploy_df_cap else cameras_grid_states

# Keep needed columns and coerce dates
cam_deploy_use <- cam_deploy_use %>%
  sf::st_drop_geometry() %>%
  mutate(
    Start_Deployment_date = suppressWarnings(as.Date(Start_Deployment_date)),
    End_Deployment_date   = suppressWarnings(as.Date(End_Deployment_date))
  )

# ---- Pull total detections from DetHist (post-filtered & collapsed exactly as for Y.o) ----
# DetHist is list over states; each element is a named list over projects, each having $detection_history
extract_cam_detections <- function(DetHist_state) {
  if (length(DetHist_state) == 0) {
    return(tibble(Project_name = character(0), TotalDetections = integer(0)))
  }
  purrr::imap_dfr(DetHist_state, function(x, project) {
    if (is.null(x) || is.null(x$detection_history)) {
      tibble(Project_name = project, TotalDetections = 0L)
    } else {
      mat <- x$detection_history
      tibble(
        Project_name = project,
        TotalDetections = sum(mat, na.rm = TRUE)
      )
    }
  })
}

# Build a state-indexed tibble of detections from DetHist
cam_det_tbl <- purrr::imap_dfr(DetHist, function(DetHist_state, state_id) {
  extract_cam_detections(DetHist_state) %>%
    mutate(State = suppressWarnings(as.integer(state_id)))
})

# Summaries per project×state from deployments
camera_table <- cam_deploy_use %>%
  filter(!is.na(Project_name), !is.na(State)) %>%
  group_by(State, Project_name) %>%
  summarise(
    `Number of stations` = n_distinct(Station_ID),
    start_min  = suppressWarnings(min(Start_Deployment_date, na.rm = TRUE)),
    end_max    = suppressWarnings(max(End_Deployment_date,   na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(`Year used` = fmt_year_used(start_min, end_max)) %>%
  dplyr::select(State, Project_name, `Number of stations`, `Year used`) %>%
  # Join detections from DetHist (truth used by Y.o)
  left_join(cam_det_tbl, by = c("State","Project_name")) %>%
  mutate(TotalDetections = dplyr::coalesce(TotalDetections, 0L)) %>%
  # Add SS[x] and Subpopulation labels
  left_join(canon_map, by = "State") %>%
  transmute(
    `Project name`  = Project_name,
    Subpopulation,
    `State Space`   = StateSpace,
    `Number of stations`,
    `Year used`,
    `Total fisher detections` = TotalDetections
  ) %>%
  arrange(Subpopulation, `State Space`, `Project name`)

cat("\n### Camera studies included (by project × state)\n")
knitr::kable(camera_table,
             caption = "Camera studies included — project × state summary (detections from DetHist / Y.o)",
             align = "l")


# =======================================================================================
# 2) DNA TABLE — project × state, consistent with DNA pipeline / Y.s construction
# =======================================================================================

# Build per-project×state totals from the post-rules DNA sf table.
# This aligns with how Y.s is built (from DNA_states after rules & state assignment).
dna_det_tbl <- DNA_states %>%
  sf::st_drop_geometry() %>%
  mutate(State = dplyr::coalesce(as.integer(State), as.integer(State.x))) %>%
  filter(!is.na(Project_name), !is.na(State)) %>%
  mutate(is_fisher = Species == species_of_interest) %>%
  group_by(State, Project_name) %>%
  summarise(
    `Total fisher detections` = sum(is_fisher, na.rm = TRUE),  # count fisher rows
    `Number of stations`      = n_distinct(Station_ID),
    # Prefer Start_date / End_date if available; otherwise try a single Date column
    start_min = suppressWarnings(min(as.Date(Start_date %||% NA), as.Date(End_date %||% NA), na.rm = TRUE)),
    end_max   = suppressWarnings(max(as.Date(End_date   %||% NA), as.Date(Start_date %||% NA), na.rm = TRUE)),
    .groups = "drop"
  )

# If Start_date/End_date are absent or NA, try fallback Date column if it exists
if (any(is.na(dna_det_tbl$start_min)) || any(is.na(dna_det_tbl$end_max))) {
  if ("Date" %in% names(sf::st_drop_geometry(DNA_states))) {
    dna_dates_alt <- DNA_states %>%
      sf::st_drop_geometry() %>%
      mutate(State = dplyr::coalesce(as.integer(State), as.integer(State.x))) %>%
      filter(!is.na(Project_name), !is.na(State)) %>%
      mutate(Date_alt = suppressWarnings(as.Date(Date))) %>%
      group_by(State, Project_name) %>%
      summarise(
        start_min_alt = min(Date_alt, na.rm = TRUE),
        end_max_alt   = max(Date_alt, na.rm = TRUE),
        .groups = "drop"
      )
    dna_det_tbl <- dna_det_tbl %>%
      left_join(dna_dates_alt, by = c("State","Project_name")) %>%
      mutate(
        start_min = ifelse(is.finite(start_min), start_min, start_min_alt),
        end_max   = ifelse(is.finite(end_max),   end_max,   end_max_alt)
      ) %>%
      dplyr::select(-start_min_alt, -end_max_alt)
  }
}

dna_table <- dna_det_tbl %>%
  mutate(`Year used` = fmt_year_used(start_min, end_max)) %>%
  left_join(canon_map, by = "State") %>%
  transmute(
    `Project name`  = Project_name,
    Subpopulation,
    `State Space`   = StateSpace,
    `Number of stations`,
    `Year used`     = dplyr::if_else(is.na(`Year used`) | `Year used` == "", "unknown", `Year used`),
    `Total fisher detections`
  ) %>%
  arrange(Subpopulation, `State Space`, `Project name`)

cat("\n### DNA studies included (by project × state)\n")
knitr::kable(dna_table,
             caption = "DNA studies included — project × state summary (aligned with Y.s pipeline)",
             align = "l")


```

## Session info
```{r session}
sessionInfo()
```
