---
title: "6 Fit Integrated Density Model (SCR + Cameras): Fully Annotated Workflow"
author: "Prepared for new analysts"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

# Overview & Purpose

This script reads the **prepared data** saved in **Script 5** and runs an **integrated density model** (DNA SCR + cameras) in **nimble**. The workflow is fully annotated for new analysts.

**Key behaviors in this script**

- **Minimal, data‑driven session reordering**: immediately after loading `data`, we reorder sessions so **camera sessions are first**, and **non‑camera sessions are last**.
- **Manual loop edit reminder**: the script prints how many sessions have/ don’t have cameras and shows exactly how to edit any loops in your **model files** or **sampler blocks** that are currently coded as `1:(N.session-1)`.
- **Choose model type** (both are linear):
  - `fixed` — fixed slopes/intercepts across sessions
  - `random` — random slopes/intercepts across sessions
  The script **sources a different Nimble model file** based on `model_type`.

> ⚠️ **Important**: We are *not* modifying your Nimble model code here. If your model file still has hard-coded loops (e.g., `for(g in 1:(N.session-1))`), use the **console message** this script prints to update them to the correct ranges based on the actual data.

---

# Parameters

```{r parameters}
# Root path for code and data
path_root <- "./3. Outputs"
#variable <- "soilMoist" # change this to whatever variable you're running (age, fcover)
#variable <- "age" 
variable <- "fcover" 

input_data_rdata <- load(file.path(path_root,  sprintf("Fit_data_Fisher_%s.RData", variable))) #output from script 5. 
# Model type: choose which Nimble model file to source
#   "fixed"  -> fixed slopes/intercepts across sessions
#   "random" -> random slopes/intercepts across sessions
model_type <- "random"   # "fixed" or "random"

# Filenames for model and helpers (adjust names if your repo differs)
fp <- function(x) file.path("./", x)

sim_single_file        <- fp("sim.SCR.Dcov.Mb.Ramsey.R")                   # single-session simulator (helper)
sim_multi_file         <- fp("sim.SCR.Dcov.Multi.Mb.Ramsey.R")             # multisession wrapper (helper)
init_multi_file        <- fp("init.SCR.Dcov.Multi.Mb.Ramsey.R")             # prepares multi-session arrays
nimble_functions_file  <- fp("NimbleFunctions SCR Dcov Multisession Mb Ramsey All Sessions.R")
sampler_functions_file <- fp("sSampler Dcov Multisession Mb Ramsey All Sessions.R")
mask_check_file        <- fp("mask.check.R")

nimble_model_file <- switch(model_type,
  fixed  = fp("NimbleModel SCR Dcov Multisession Mb Ramsey Marginal All Sessions_fixed.R"),
  random = fp("NimbleModel SCR Dcov Multisession Mb Ramsey Marginal All Sessions.R"),
  stop("model_type must be 'fixed' or 'random'"))

# Data augmentation per session (length N.session). Adjust as needed
M_default <- c(125, 250, 350, 400, 550, 250)

# MCMC settings
n_iter   <- 60000
nburnin  <- 20000
n_chains <- 3
thin_par <- 10
thin_aux <- 25

# Output

run_date <- format(Sys.Date(), "%Y-%m-%d")
out_summary_csv <- sprintf("new_mcmc_summary_%s_%s_%s.csv",
                           model_type, n_iter, run_date)
out_trace_pdf   <- sprintf("new_mcmc_trace_%s_%s_%s.pdf",
                           model_type, n_iter, run_date)

```

# Setup

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 6)

# Packages
suppressPackageStartupMessages({
  library(nimble)
  library(coda)
  library(RColorBrewer)
  library(MCMCvis)
})

# Nimble options (version-dependent)
nimbleOptions(determinePredictiveNodesInModel = FALSE)
# For nimble < 0.13.1, you might need instead:
# nimble:::setNimbleOption('MCMCjointlySamplePredictiveBranches', FALSE)
```

# Source model and helpers

```{r source-files}
source(sim_single_file)
source(sim_multi_file)
source(init_multi_file)
source(nimble_model_file)      # defines NimModel (fixed or random)
source(nimble_functions_file)
source(sampler_functions_file)
source(mask_check_file)
```

# Load data and **minimally** reorder sessions (camera-first)

```{r load-and-reorder}
# Read the Script-5 output (named 'data')
#load(input_data_rdata)
stopifnot(exists("data"))
N.session <- length(data)
cat("N.session (before reorder):", N.session, "\n")

# Minimal reorder: put camera sessions first, no-camera sessions last
has_cams_vec <- vapply(seq_len(N.session), function(g) {
  x2_ok <- !is.null(data[[g]]$X2) && is.matrix(data[[g]]$X2) && nrow(data[[g]]$X2) > 0
  y2_ok <- !is.null(data[[g]]$y2) && is.matrix(data[[g]]$y2) && all(dim(data[[g]]$y2) > 0)
  x2_ok && y2_ok
}, logical(1))

sessions_with_cams    <- which(has_cams_vec)
sessions_without_cams <- which(!has_cams_vec)

new_order <- c(sessions_with_cams, sessions_without_cams)
data <- data[new_order]

N_cam    <- length(sessions_with_cams)
N_no_cam <- length(sessions_without_cams)

cat("\n[Reorder] New session order (camera-first): ", paste(new_order, collapse = ", "),
    "\n[Reorder] N_cam = ", N_cam, " | N_no_cam = ", N_no_cam,
    "\n[Reorder] Manual edit needed in model/samplers:",
    "\n          • Change camera loops from  for(g in 1:(N.session-1))  to  for(g in 1:", N_cam, ")",
    if (N_no_cam > 0)
      paste0("\n          • No-camera loops should target the last ", N_no_cam,
             " session(s): for(g in ", (N.session - N_no_cam + 1), ":", N.session, ")")
    else "\n          • No-camera loops can be removed (none present).",
    "\n\n", sep = "")
```

# Mask sanity check (per session)

```{r mask-check}
for (g in seq_len(N.session)) {
  mask.check(dSS = data[[g]]$dSS,
             cells = data[[g]]$cells,
             n.cells = data[[g]]$n.cells,
             n.cells.x = data[[g]]$n.cells.x,
             n.cells.y = data[[g]]$n.cells.y,
             res = data[[g]]$res,
             xlim = data[[g]]$xlim,
             ylim = data[[g]]$ylim,
             x.vals = data[[g]]$x.vals,
             y.vals = data[[g]]$y.vals)
}
```

# Standardize the density covariate (pooled across sessions)

```{r standardize-dcov}
D.cov.all.vals <- unlist(lapply(data, function(x) x$D.cov))
mean.D.cov <- mean(D.cov.all.vals)
sd.D.cov   <- sd(D.cov.all.vals)
for (g in seq_len(N.session)) data[[g]]$D.cov <- (data[[g]]$D.cov - mean.D.cov) / sd.D.cov
```

# Initialize multisession structures

```{r init-multi}
# M vector follows the (reordered) sessions
M <- if (length(M_default) == 1L) rep(M_default, N.session) else M_default
stopifnot(length(M) == N.session)

nimbuild <- init.SCR.Dcov.Multi.Mb.Ramsey(data, M)

# Quick visual QA: initialized activity centers are in habitat
par(mfrow = c(2, 3))
for (g in seq_len(N.session)) {
  image(data[[g]]$x.vals, data[[g]]$y.vals,
        matrix(data[[g]]$InSS, data[[g]]$n.cells.x, data[[g]]$n.cells.y),
        main = sprintf("Init s in InSS (g=%d)", g), col = hcl.colors(12, "Greens", rev = TRUE))
  points(nimbuild$s.init[g,,], pch = 16, cex = 0.6)
}
par(mfrow = c(1, 1))
```

# Build inits, constants, data

```{r nimble-objects}
# Inits (must include z.init, N.init, and s.init for DA scheme)
Niminits <- list(
  z = nimbuild$z.init,
  N = rowSums(nimbuild$z.init),
  s = nimbuild$s.init,
  beta0.p0.SCR.fixed = -3,
  beta1.p0.SCR.fixed = 3,
  p0.USCR = rep(0.1, N.session),
  sigma.fixed = 4.5,
  D0 = rowSums(nimbuild$z.init) / (rowSums(nimbuild$InSS) * nimbuild$cellArea),
  D.beta1 = rep(0, N.session)
)

# Constants (match Nimble model spec)
constants <- list(
  N.session = N.session,
  M         = M,
  J1        = nimbuild$J1,
  J2        = nimbuild$J2,
  K1        = nimbuild$K1,
  K2D1      = nimbuild$K2D1,
  K1D2      = nimbuild$K1D2,
  xlim      = nimbuild$xlim,
  ylim      = nimbuild$ylim,
  D.cov     = nimbuild$D.cov,
  cellArea  = nimbuild$cellArea,
  n.cells   = nimbuild$n.cells,
  res       = nimbuild$res
)

# Data
Nimdata <- list(
  y1        = nimbuild$y1,
  y1.state  = nimbuild$y1.state,
  y2        = nimbuild$y2,
  z         = nimbuild$z.data,
  X1        = nimbuild$X1,
  X2        = nimbuild$X2,
  dummy.data= nimbuild$dummy.data,
  cells     = nimbuild$cells,
  InSS      = nimbuild$InSS
)
```

# Configure samplers and monitors

```{r mcmc-config}
# Monitors (both model types are linear; difference is fixed vs random effects in the sourced file)
parameters  <- c('N','lambda.N','beta0.p0.SCR.fixed','beta1.p0.SCR.fixed','p0.USCR','sigma.fixed','D0','D.beta1')
parameters2 <- c('lambda.cell','s.cell','D0')   # record D0 for plotting/diagnostics
nt  <- thin_par
nt2 <- thin_aux

# Build model (ignore version-dependent warnings; see header notes)
Rmodel <- nimbleModel(code = NimModel, constants = constants, data = Nimdata, check = FALSE, inits = Niminits)
config.nodes <- c('beta0.p0.SCR.fixed','beta1.p0.SCR.fixed','p0.USCR','sigma.fixed','D0','D.beta1')
conf <- configureMCMC(Rmodel, monitors = parameters, thin = nt, monitors2 = parameters2, thin2 = nt2,
                      nodes = config.nodes, useConjugacy = FALSE)

# Required sampler replacements (camera-aware vs no-camera)
z_ups <- round(M * 0.25)

# Camera sessions (FIRST N_cam after reorder)
if (N_cam > 0) {
  for (g in 1:N_cam) {
    conf$addSampler(target = paste0("N[", g, "]"), type = 'zSamplerCams',
                    control = list(g = g, inds.detected = 1:nimbuild$n1[g], z.ups = z_ups[g],
                                   J1 = nimbuild$J1[g], J2 = nimbuild$J2[g], K1 = nimbuild$K1[g], M = M[g]),
                    silent = TRUE)
    for (i in seq_len(M[g])) {
      conf$addSampler(target = paste0("s[", g, ",", i, ", 1:2]"), type = 'sSamplerCams',
                      control = list(g = g, i = i, J1 = nimbuild$J1[g], J2 = nimbuild$J2[g], K1 = nimbuild$K1[g],
                                     res = nimbuild$res[g], n.cells = nimbuild$n.cells[g], n.cells.x = nimbuild$n.cells.x[g],
                                     n.cells.y = nimbuild$n.cells.y[g], xlim = nimbuild$xlim[g,], ylim = nimbuild$ylim[g,]),
                      silent = TRUE)
    }
  }
}

# No-camera sessions (LAST N_no_cam after reorder)
if (N_no_cam > 0) {
  for (g in (N.session - N_no_cam + 1):N.session) {
    conf$addSampler(target = paste0("N[", g, "]"), type = 'zSamplerNoCams',
                    control = list(g = g, inds.detected = 1:nimbuild$n1[g], z.ups = z_ups[g],
                                   J1 = nimbuild$J1[g], K1 = nimbuild$K1[g], M = M[g]),
                    silent = TRUE)
    for (i in seq_len(M[g])) {
      conf$addSampler(target = paste0("s[", g, ",", i, ", 1:2]"), type = 'sSamplerNoCams',
                      control = list(g = g, i = i, J1 = nimbuild$J1[g], K1 = nimbuild$K1[g],
                                     res = nimbuild$res[g], n.cells = nimbuild$n.cells[g], n.cells.x = nimbuild$n.cells.x[g],
                                     n.cells.y = nimbuild$n.cells.y[g], xlim = nimbuild$xlim[g,], ylim = nimbuild$ylim[g,]),
                      silent = TRUE)
    }
  }
}

# Often helpful to block these parameters per session
for (g in seq_len(N.session)) {
  conf$addSampler(target = c(paste0("D0[", g, "]"), paste0("D.beta1[", g, "]")),
                  type = 'AF_slice', control = list(adaptive = TRUE), silent = TRUE)
}
```

# Compile

```{r run}
Rmcmc  <- buildMCMC(conf)
Cmodel <- compileNimble(Rmodel)
Cmcmc  <- compileNimble(Rmcmc, project = Rmodel)

```

# Run 

Loop to run a chain, save and run next chain. This helps when problems when no enough memory. By setting a seed you can replicate posteriors. 

```{r}
start.time <- Sys.time()
for (ch in 1:n_chains) {
  set.seed(123 + ch)  # different seed per chain for independence
  # Run ONE chain at a time
  samp <- runMCMC(Cmcmc,
                  niter             = n_iter,
                  nburnin           = nburnin,
                  nchains           = 1, ## always 1 as n_chains is already set in the loop
                  samplesAsCodaMCMC = TRUE)  # returns an mcmc.list of length 1
  # Save as RDS (preserves types cleanly)
  saveRDS(samp, file = file.path(path_root, sprintf("mcmc_chain_%s_%02d.rds", variable, ch)))
  # Optionally also save as CSV for quick inspection
  m <- as.matrix(samp[[1]])
  write.csv(m, file = file.path(path_root, sprintf("mcmc_chain_%s_%02d.csv", variable, ch)), row.names = TRUE)
  rm(samp, m); gc()  # keep memory tidy for long runs
}

end.time <- Sys.time()
cat("Total wall time:", end.time - start.time, "\n")

```

# Summarize and export

```{r summarize, message=FALSE}

# MCMC settings
#n_iter   <- 60000
n_chains <- 3

# Output
model_type <-  "random"
run_date <- format(Sys.Date(), "%Y-%m-%d")
out_summary_csv <- file.path(path_root, sprintf("mcmc_summary_%s_%s_%s.csv",
                           variable, model_type, n_iter, run_date))
out_trace_pdf   <- sprintf("mcmc_trace_%s_%s_%s.pdf",
                           variable, model_type, n_iter, run_date)


# 1) Read the saved chains (each file contains an mcmc.list of length 1)
files <- file.path(path_root, sprintf("mcmc_chain_%s_%02d.rds", variable, ch))
stopifnot(all(file.exists(files)))
chains_raw <- lapply(files, readRDS)  # each is mcmc.list of length 1
 
# 2) Flatten to a single mcmc.list with one chain per element
mcmc_list <- do.call(mcmc.list, lapply(chains_raw, `[[`, 1))
 
# ---- If you want to drop parameters BEFORE passing to MCMCvis: ----
# Convert to matrices -> drop columns -> convert back to mcmc.list
 
# Find which columns to drop using the first chain as template
mat1 <- as.matrix(mcmc_list[[1]])
drop <- colnames(mat1) %in% c(paste0("lambda.N[", 1:6, "]"), c("p0.USCR[6]"))  #may need to add "p0.USCR[x]" for other sites without cameras
 
# Apply the same filter to all chains (safely handling missing names)
mcmc_list_filtered <- mcmc.list(lapply(mcmc_list, function(ch) {
  m <- as.matrix(ch)
  # align columns to the template names to avoid subtle ordering differences
  m <- m[, colnames(mat1), drop = FALSE]
  m <- m[, !drop, drop = FALSE]
  as.mcmc(m)
}))
 
# 3) Summaries and trace plots with MCMCvis
dat <- MCMCsummary(mcmc_list_filtered, round = 2)
write.csv(dat, out_summary_csv, row.names = TRUE)
 
MCMCtrace(object  = mcmc_list_filtered,
          pdf     = TRUE,
          filename= out_trace_pdf,
          ind     = TRUE,   # different lines per chain
          Rhat    = TRUE,   # add Rhat
          n.eff   = TRUE,   # add effective sample size
          params  = "all")

#If you get an error similar to this:Error in quantile.default(newX[, i], ...) : 
#  missing values and NaN's not allowed if 'na.rm' is FALSE
# check the drop object, you might need to include other p0.USCR[6] if you are missing cameras in state spaces

```

# Diagnostics: final activity centers in habitat

```{r final-centers}
par(mfrow = c(2, 3))
for (g in seq_len(N.session)) {
  image(data[[g]]$x.vals, data[[g]]$y.vals,
        matrix(data[[g]]$InSS, data[[g]]$n.cells.x, data[[g]]$n.cells.y),
        main = sprintf("Final s in InSS (g=%d)", g), col = hcl.colors(12, "Greens", rev = TRUE))
  points(Cmodel$s[g,,], pch = 16, cex = 0.6)
}
par(mfrow = c(1, 1))
```

# Notes on warnings & camera logic

- You may see version-dependent warnings at model definition time about dimensions for `p0.USCR`, `y2`, `X2`, `cells`. These typically reflect **mismatches between data order and model loop assumptions**. We **reordered sessions camera-first** to align with models that expect the first *K* sessions to have cameras.
- This script prints **`N_cam`** and **`N_no_cam`** after loading data so you can **manually edit** any hard-coded camera loop bounds (e.g., change `for(g in 1:(N.session-1))` to `for(g in 1:N_cam)`).
- Use `model_type <- "fixed"` or `"random"` at the top to choose the Nimble model file to source.

# Session Info

```{r session-info}
sessionInfo()
```
